High-level cognition is supported by at least second order dynamic correlations in neural activity patterns

### 2019-09-05 paper/main.tex
spell checked document

To gain insights into the above aspects of the neural code, we
developed a computational framework for estimating dynamic high-order
correlations in timeseries data. This framework provides an important
advance, in that it enables us to examine patterns in higher-order
correlations that are computationally intractable to estimate via
conventional methods.  Given a multivariate timeseries, our framework
provides timepoint-by-timepoint estimates of the first-order
correlations, second-order correlations, and so on (up to tenth-order
correlations in this manuscript).  Our approach combines a
kernel-based method for computing dynamic correlations in timeseries
data with a dimensionality reduction step (Fig.~\ref{fig:patterns}b)
that projects the resulting dynamic correlations into a
low-dimensional space.  We explored two dimensionality reduction
approaches: principle components analysis~\citep[PCA;][]{Pear01},
which preserves an approximately invertable transformation back to the
original data~\citep[e.g., this follows related approaches taken
by][]{McInJirs19, TokeSomm19, GonzEtal19}; and a second non-invertible
algorithm that explored patterns in eigenvector
centrality~\citep{Land95}.  This latter approach characterizes
correlations between each feature dimension's relative
\textit{position} in the network in favor of the specific activity
histories of different features~\citep[also see][]{BetzEtal19,
  SizeEtal18, ReimEtal17}.

We validated our approach using synthetic data where the underlying
correlations were known.  We then applied our framework to a
neuroimaging dataset collected as participants
listened to either an audio recording of a ten-minute story or a
temporally scrambled version of the story, or underwent a resting
state scan~\citep{SimoEtal16}.  We used a subset of the data to train
across-participant classifiers to decode listening times using a blend
of neural features (comprising neural activity patterns, as well as
different orders of correlations between those patterns that were
inferred using our computational framework).  We found that both the
PCA-based and eigenvector centrality-based approaches yielded neural
patterns that could be used to decode accurately.  Both approaches
also yielded the best decoding accuracy for data collected during
(intact) story listening when high-order (PCA: second-order;
eigenvector centrality: fourth-order) dynamic correlation patterns
were included as features.  When we trained classifiers on the
scambled stories or resting state data, only lower-order dynamic
patterns were informative to the decoders.  Taken together, our
results indicate that high-level cognition is supported by high-order
dynamic patterns of communication between brain structures.

\section*{Methods}
Our general approach to comprises four general steps
(Fig.~\ref{fig:methods}).  First, we derive a kernel-based approach to
computing dynamic pairwise correlations in a $T$ (timepoints) by $K$
(features) multivariate timeseries, $\mathbf{X}_0$.  This yields a $T$
by $\mathcal{O}(K^2)$ matrix of dynamic correlations, $\mathbf{Y}_1$,
where each row comprises the upper triangle of the correlation matrix
at a single timepoint, reshaped into a row vector (this reshaped
vector is $(\frac{K^2 - K}{2})$-dimensional).  Second, we apply a
dimensionality reduction step to project the matrix of dynamic
correlations back onto a $K$-dimensional space.  This yields a $T$ by
$K$ matrix, $\mathbf{X}_1$, that reflects an approximation of the
dynamic correlations reflected in the original data.  Third, we use
repeated applications of the kernel-based dynamic correlation step to
$\mathbf{X}_n$ and the dimensionality reduction step to the resulting
$\mathbf{Y}_{n+1}$ to estimate high-order dynamnic correlations.  Each
application of these steps to a $T$ by $K$ time series $\mathbf{X}_n$
yields a $T$ by $K$ matrix, $\mathbf{X}_{n+1}$, that reflects the
dynamic correlations between the columns of $\mathbf{X}_n$.  In this
way, we refer to $n$ as the \textit{order} of the timeseries, where
$\mathbf{X}_0$ (order 0) denotes the original data and $\mathbf{X}_n$
denotes $n^\mathrm{th}$-order dynamic correlations between the columns
of $\mathbf{X}_0$.  Finally, we use a cross-validation--based decoding
approach to evaluate how well information contained in a given order
(or weighted mixture of orders) may be used to decode relevant
cognitive states.  If including a given $\mathbf{X}_n$ in the feature
set yields higher classification accuracy on held-out data, we
interpret this as evidence that the given cognitive states are
reflected in patterns of $n^\mathrm{th}$-order correlations.  All of
the code used to produce the figures and results in this manuscript,
along with links to the corresponding datasets, may be found at
\href{https://github.com/ContextLab/timecorr-paper}{github.com/ContextLab/timecorr-paper}.  In addition, we have
released a Python toolbox for computing dynamic high-order
correlations in timeseries data; our toolbox may be found at
\href{https://timecorr.readthedocs.io/}{timecorr.readthedocs.io}. \textbf{JRM NOTE: CHECK LINK}

\begin{figure}
  \centering
  \includegraphics[width=0.4\textwidth]{figs/methods_fig}
  \caption{\textbf{Estimating dynamnic high-order correlations.}  Given
    a $T$ by $K$ matrix of multivariate timeseries data,
    $\mathbf{Y}_n$ (where $n \in \mathbb{N}, n \geq 0$), we use
    Equation~\ref{eqn:timecorr} to compute a timeseries of
    $K$ by $K$ correlation matrices, $\mathbf{Y}_{n+1}$.  We then
    approximate $\mathbf{Y}_{n+1}$ with the $T$ by $K$ matrix
    $\mathbf{X}_{n+1}$.  This process may be repeated to scalably estimate
    iteratively higher-order correlations in the data.  Note that the
    transposes of $\mathbf{Y}_n$ and $\mathbf{X}_{n+1}$ are displayed
    in the figure for compactness.
  \label{fig:methods}}
\end{figure}

@@

The modern library of dimensionality reduction algorithms include
Principal Components Analysis~\citep[PCA; ][]{Pear01}, Probabilistic
PCA~\citep[PPCA; ][]{TippBish99}, Exploratory Factor
Analysis~\citep[EFA; ][]{Spea04}, Independent Components
Analysis~\citep[ICA; ][]{JuttHera91, ComoEtal91}, $t$-Stochastic
Neighbor Embedding~\citep[$t$-SNE; ][]{MaatHint08}, Uniform Manifold
Approximation and Projection~\citep[UMAP; ][]{McInHeal18},
non-negative matrix factorization~\citep[NMF; ][]{LeeSeun99},
Topographic Factor Analysis (TFA)~\cite{MannEtal14b}, Hierarchical
Topographic Factor analysis (HTFA)~\cite{MannEtal18}, Topographic
Latent Source Analysis (TLSA)~\cite{GersEtal11}, Dictionary
learning~\citep{MairEtal09a, MairEtal09b}, deep
autoencoders~\citep{HintSala06}, among others.  While complete
characterizations of each of these algorithms is beyond the scope of
the present manuscript, the general intuition driving these approaches
is to compute the $T$ by $I$ matrix, $\mathbf{X}$, that is closest
to the original $T$ by $J$ matrix, $\mathbf{Y}$, where (typically)
$I \ll J$.  The different approaches place different constraints on
what properties $\mathbf{X}$ must satisfy and which aspects of
the data are compared (and how) to characterize the match between
$\mathbf{X}$ and $\mathbf{Y}$.

Applying dimensionality reduction algorithms to $\mathbf{Y}$ yields a
$\mathbf{X}$ whose columns reflect weighted combinations (or nonlinear
transformations) of the original columns of $\mathbf{Y}$.  This has
two main consequences.  First, with each repeated dimensionality
reduction, the resulting $\mathbf{X}_n$ has lower and lower fidelity
(with respect to what the ``true'' $\mathbf{Y}_n$ might have looked
like without using dimensionality reduction to maintain scalability).
In other words, computing $\mathbf{X}_n$ is a lossy operation.
Second, whereas each columns of $\mathbf{Y}_n$ may always be mapped
directly onto specific pairs of columns of $\mathbf{Y}_{n-1}$, the
columns of $\mathbf{X}_n$ reflect weighted combinations and/or
nonlinear transformations of the columns of $\mathbf{Y}_n$.  Many
dimensionality reduction algorithms are invertable (or approximately
invertable).  However, attempting to map a given $\mathbf{X}_n$ back
onto the original feature space of $\mathbf{Y}_0$ will usually require
$\mathcal{O}(TK^{2n})$ space and therefore quickly becomes intractable
as $n$ or $K$ grow large.

@@

In general, graph measures take as input a matrix of
interactions (e.g., using the above notation, an $K$ by $K$
correlation matrix or binarized correlation matrix reconstituded from
a single timepoint's row of $\mathbf{Y}$) and return as output a set
of $K$ measures describing how each node (feature) sits within that
correlation matrix with respect to the rest of the population.  Widely
used measures include betweeness centrality~\citep[the proportion of
shortest paths between each pair of nodes in the population that
involves the given node in question; e.g., ][]{Newm05, OpsaEtal10,
  Bart04, GeisEtal08, Free77}; diversity and
dissimilarity~\citep[characterizations of how differently connected a
given node is from others in the population; e.g., ][]{Rao82, Lin09,
  RicoSzei06}; Eigenvector centrality and pagerank
centrality~\citep[measures of how influential a given node is within
the broader network; e.g., ][]{Newm08, Bona07, LohmEtal10,
  HaluEtal13}; transfer entropy and flow coefficients~\citep[a measure
of how much information is flowing from a given node to other nodes in
the network; e.g., ][]{HoneEtal07, Schr00}; $k$-coreness
centrality~\citep[a measure of the connectivity of a node within its
local sub-graph; e.g., ][]{AlvaEtal05, ChriFowl10}; within-module
degree~\citep[a measure of how many connections a node has to its
close neighbors in the network; e.g., ][]{RubiSpor10}; participation
coefficient~\citep[a measure of the diversity of a node's connections
to different sub-graphs in the network; e.g., ][]{RubiSpor10}; and
sub-graph centrality~\citep[a measure of a node's participation in all
of the network's sub-graphs; e.g., ][]{EstrRodr05}; among others.

@@

\paragraph*{Event.} We generated a fourth set of 10 synthetic datasets
whose underlying correlation matrices exhibited prolonged intervals of
stability, intersperced with abrupt changes.  For each dataset, we
used Equation~\ref{eqn:randcorr} to generate 5 random covariance
matrices.  We constructed a timeseries where each set of 60 consecutive samples
was drawn from a Gaussian with the same covariance matrix.  These
datasets were intended to simulate a system that undergoes occasional
abrupt state changes.

\subsubsection*{Functional neuroimaging data collected during story
  listening}
We examined an fMRI dataset collected by \cite{SimoEtal16} that the
authors have made publically available at
\href{http://arks.princeton.edu/ark:/88435/dsp015d86p269k}{arks.princeton.edu/ark:/88435/dsp015d86p269k}.  The dataset
comprises neuroimaging data collected as participants listened
to an audio recording of a story (intact condition; 36 participants),
listened to time scrambled recordings of the same story (17
participants in the paragraph-scrambled condition listened to the
paragraphs in a randomized order and 36 in the word-scrambled
condition listened to the words in a randomized order), or lay resting
with their eyes open in the scanner (rest condition; 36
participants).  Full neuroimaging details may be found in the original
paper for which the data were collected~\citep{SimoEtal16}.

@@

\subsubsection*{Identifying robust decoding results}
The temporal decoding procedure we use to estimate which neural
features support ongoing cognitive processing is goverened by many
parameters.  For example, Equation~\ref{eqn:timecorr} requires
defining a kernel function, which can take on different shapes and
widths.  For a fixed set of neural features, each of these parameters
can yield different decoding accuracies.  Further, the best decoding
accuracy for a given timepoint may be reliably achieved by one set of
parameters, whereas the best decoding accuracy for another timepoint
might be reliably achieved by a different set of parameters, and the
best decoding accuracy across \textit{all} timepoints might be
reliably achieved by still another different set of parameters.
Rather than attempting to maximize decoding accuracy, we sought to
discover the trends in the data that were robust to specific
classifier parameters choices.  Specifically, we sought to
characterize how decoding accuracy varied (under different
experimental conditions) as a function of which neural features were
considered.

@@

\section*{Results}
We sought to understand whether high-level cognition is supported by
dynamic patterns of high-order correlations.  To that end, we
developed a computational framework for estimating the dynamics of
high-order correlations in multivariate timeseries data (see
\textit{Dynamic inter-subject functional connectivity (DISFC)} and
\textit{Dynamic higher-order correlations}).  We evaluated the
efficacy of this framework at recovering known patterns in several
synthetic datasets (see \textit{Synthetic data}).  We then applied the
framework to a public fMRI dataset collected as participants listened
to an auditorally presented story, a temporally scrambled version of
the story, or underwent a resting state scan (see \textit{Functional
  neuroimaing data collected during story listening}).  We used the
relative decoding accuracies of classifiers trained on
different sets of neural features to estimate which types of features
reflected ongoing cognitive processing.

@@

\begin{figure}[tp]
  \centering
  \includegraphics[width=\textwidth]{figs/decode_level}
  \caption{\textbf{Across-participant decoding accuracy varies with
      correlation order and cognitive engagement.}
    \textbf{a. Decoding accuracy as a function of order: PCA.}
    \textit{Order} ($x$-axis) refers to the maximum order of dynamic
    correlations that were available to the classifiers (see
    \textit{Feature weighting and testing}).  The reported
    across-participant decoding accuracies are averaged over all
    kernel shapes and widths (see \textit{Identifying robust decoding
      results}).  The $y$-values are displayed relative to chance
    accuracy (intact: $\frac{1}{300}$; paragraph: $\frac{1}{272}$;
    word: $\frac{1}{300}$; rest: $\frac{1}{400}$).  The error bars
    denote 95\% confidence intervals across cross-validation folds
    (i.e., random assignments of participants to the training and test
    sets).  The colors denote the experimental condition.  Arrows
    denote sets of features that yielded reliably higher (upwards
    facing) or lower (downward facing) decoding accuracy than the mean
    of all other features (via a two-tailed test, thresholded at
    $p < 0.05$).  The circled values represent the maximum decoding
    accuracy within each experimental condition.
    \textbf{b. Normalized decoding accuracy as a function of order:
      PCA.}  This panel displays the same results as Panel a, but here
    each curve has been normalized to have a maximum value of 1 and
    and a minimum value of 0 (including the upper and lower bounds of
    the respective 95\% confidence intervals).  Panels a and b used
    PCA to project each high-dimensional pattern of dynamic
    correlations onto a lower-dimensional space.  \textbf{c.  Decoding
      accuracy as a function of order: eigenvector centrality.} This
    panel is in the same format as Panel a, but here eigenvector
    centrality has been used to project the high-dimensional patterns
    of dynamic correlations onto a lower-dimensional space.
    \textbf{d. Noramlized decoding accuracy as a function of order:
      eigenvector centrality.} This panel is in the same format as
    Panel b, but here eigenvector centrality has been used to project
    the high-dimensional patterns of dynamic correlations onto a
    lower-dimensional space.}
  \label{fig:decoding}
\end{figure}

Both sets of temporal decoding analyses yielded qualitatively similar
results for the auditory (non-rest) conditions of the experiment
(Fig.~\ref{fig:decoding}: pink, yellow, and teal lines).  The highest
decoding accuracy for participants who listened to the intact
(unscrambled) story was achieved using high-order dynamic correlations
(PCA: second-order; eigenvector-centrality: fourth-order).  Scambled
versions of the story were best decoded by lower-order correlations
(PCA/paragraph: first-order; PCA/word: order zero; eigenvector
centrality/paragraph: order zero; eigenvector centrality/word: order
zero).  The two sets of analyses yielded different decoding results on
resting state data (Fig.~\ref{fig:decoding}: purple lines).  We note
could be decoded only very slightly above chance; we speculate that
the decoders might have picked up on attentional drift, boredom, or
tiredness; we hypothesize these all increase throughought the resting
state scan.  Our decoders might be picking up on aspects of these
loosely defined cognitive states that are common across individuals.
The PCA-based approach achieved the highest resting state decoding
accuracy using order zero features (non-correlational
activation-based), whereas the eigenvector centrality-based approach
achieved the highest resting state decoding accuracy using
second-order correlations.  Taken together, these analyses indicate
that high-level cognitive processing (while listening to the intact
story) is reflected in the dynamics of high-order correlations in
brain activity, whereas lower-level cognitive processing (while
listening to scrambled versions of the story that lack rich meaning)
is reflected in the dynamics of lower-order correlations and
non-correlational activity dynamics.  Further, these patterns are
associated both with the underlying activity patterns (characterized
using PCA) and also with the changing relative positions that
different brain areas occupy in their associated networks
(characterized using eigenvector centrality).

@@

The notion that cognition is reflected in (and possibly mediated by)
patterns of network dynamics has been suggested by or proposed in
myriad empirical studies and reviews~\citep[e.g., ][]{DemeEtal19,
  Turk13, LuriEtal18, FongEtal19, ParkEtal18b, PretEtal17,
  RoyEtal19, LiegEtal19, ZouEtal19, ChanGlov10, GonzEtal19}.  Our
study extends this line of work by finding cognitively relevant
higher-order network dynamics that reflect ongoing cognition.  Our
findings complement other work that uses graph theory and topology to
characterize how brain networks reconifigure during
cognition~\citep[e.g., ][]{BassEtal06, ZhenEtal19, McInJirs19,
  TokeSomm19, SizeEtal18, ReimEtal17, BetzEtal19}.

>>>>

To gain insights into the above aspects of the neural code, we
developed a computational framework for estimating dynamic high-order
correlations in timeseries data. This framework provides an important
advance, in that it enables us to examine patterns in higher-order
correlations that are computationally intractable to estimate via
conventional methods.  Given a multivariate timeseries, our framework
provides timepoint-by-timepoint estimates of the first-order
correlations, second-order correlations, and so on (up to tenth-order
correlations in this manuscript).  Our approach combines a
kernel-based method for computing dynamic correlations in timeseries
data with a dimensionality reduction step (Fig.~\ref{fig:patterns}b)
that projects the resulting dynamic correlations into a
low-dimensional space.  We explored two dimensionality reduction
approaches: principle components analysis~\citep[PCA;][]{Pear01},
which preserves an approximately invertible transformation back to the
original data~\citep[e.g., this follows related approaches taken
by][]{McInJirs19, TokeSomm19, GonzEtal19}; and a second non-invertible
algorithm that explored patterns in eigenvector
centrality~\citep{Land95}.  This latter approach characterizes
correlations between each feature dimension's relative
\textit{position} in the network in favor of the specific activity
histories of different features~\citep[also see][]{BetzEtal19,
  SizeEtal18, ReimEtal17}.

We validated our approach using synthetic data where the underlying
correlations were known.  We then applied our framework to a
neuroimaging dataset collected as participants
listened to either an audio recording of a ten-minute story or a
temporally scrambled version of the story, or underwent a resting
state scan~\citep{SimoEtal16}.  We used a subset of the data to train
across-participant classifiers to decode listening times using a blend
of neural features (comprising neural activity patterns, as well as
different orders of correlations between those patterns that were
inferred using our computational framework).  We found that both the
PCA-based and eigenvector centrality-based approaches yielded neural
patterns that could be used to decode accurately.  Both approaches
also yielded the best decoding accuracy for data collected during
(intact) story listening when high-order (PCA: second-order;
eigenvector centrality: fourth-order) dynamic correlation patterns
were included as features.  When we trained classifiers on the
scrambled stories or resting state data, only lower-order dynamic
patterns were informative to the decoders.  Taken together, our
results indicate that high-level cognition is supported by high-order
dynamic patterns of communication between brain structures.

\section*{Methods}
Our general approach to comprises four general steps
(Fig.~\ref{fig:methods}).  First, we derive a kernel-based approach to
computing dynamic pairwise correlations in a $T$ (timepoints) by $K$
(features) multivariate timeseries, $\mathbf{X}_0$.  This yields a $T$
by $\mathcal{O}(K^2)$ matrix of dynamic correlations, $\mathbf{Y}_1$,
where each row comprises the upper triangle of the correlation matrix
at a single timepoint, reshaped into a row vector (this reshaped
vector is $(\frac{K^2 - K}{2})$-dimensional).  Second, we apply a
dimensionality reduction step to project the matrix of dynamic
correlations back onto a $K$-dimensional space.  This yields a $T$ by
$K$ matrix, $\mathbf{X}_1$, that reflects an approximation of the
dynamic correlations reflected in the original data.  Third, we use
repeated applications of the kernel-based dynamic correlation step to
$\mathbf{X}_n$ and the dimensionality reduction step to the resulting
$\mathbf{Y}_{n+1}$ to estimate high-order dynamic correlations.  Each
application of these steps to a $T$ by $K$ time series $\mathbf{X}_n$
yields a $T$ by $K$ matrix, $\mathbf{X}_{n+1}$, that reflects the
dynamic correlations between the columns of $\mathbf{X}_n$.  In this
way, we refer to $n$ as the \textit{order} of the timeseries, where
$\mathbf{X}_0$ (order 0) denotes the original data and $\mathbf{X}_n$
denotes $n^\mathrm{th}$-order dynamic correlations between the columns
of $\mathbf{X}_0$.  Finally, we use a cross-validation--based decoding
approach to evaluate how well information contained in a given order
(or weighted mixture of orders) may be used to decode relevant
cognitive states.  If including a given $\mathbf{X}_n$ in the feature
set yields higher classification accuracy on held-out data, we
interpret this as evidence that the given cognitive states are
reflected in patterns of $n^\mathrm{th}$-order correlations.  All of
the code used to produce the figures and results in this manuscript,
along with links to the corresponding datasets, may be found at
\href{https://github.com/ContextLab/timecorr-paper}{github.com/ContextLab/timecorr-paper}.  In addition, we have
released a Python toolbox for computing dynamic high-order
correlations in timeseries data; our toolbox may be found at
\href{https://timecorr.readthedocs.io/}{timecorr.readthedocs.io}. \textbf{JRM NOTE: CHECK LINK}

\begin{figure}
  \centering
  \includegraphics[width=0.4\textwidth]{figs/methods_fig}
  \caption{\textbf{Estimating dynamic high-order correlations.}  Given
    a $T$ by $K$ matrix of multivariate timeseries data,
    $\mathbf{Y}_n$ (where $n \in \mathbb{N}, n \geq 0$), we use
    Equation~\ref{eqn:timecorr} to compute a timeseries of
    $K$ by $K$ correlation matrices, $\mathbf{Y}_{n+1}$.  We then
    approximate $\mathbf{Y}_{n+1}$ with the $T$ by $K$ matrix
    $\mathbf{X}_{n+1}$.  This process may be repeated to scalably estimate
    iteratively higher-order correlations in the data.  Note that the
    transposes of $\mathbf{Y}_n$ and $\mathbf{X}_{n+1}$ are displayed
    in the figure for compactness.
  \label{fig:methods}}
\end{figure}

@@

The modern library of dimensionality reduction algorithms include
Principal Components Analysis~\citep[PCA; ][]{Pear01}, Probabilistic
PCA~\citep[PPCA; ][]{TippBish99}, Exploratory Factor
Analysis~\citep[EFA; ][]{Spea04}, Independent Components
Analysis~\citep[ICA; ][]{JuttHera91, ComoEtal91}, $t$-Stochastic
Neighbor Embedding~\citep[$t$-SNE; ][]{MaatHint08}, Uniform Manifold
Approximation and Projection~\citep[UMAP; ][]{McInHeal18},
non-negative matrix factorization~\citep[NMF; ][]{LeeSeun99},
Topographic Factor Analysis (TFA)~\cite{MannEtal14b}, Hierarchical
Topographic Factor analysis (HTFA)~\cite{MannEtal18}, Topographic
Latent Source Analysis (TLSA)~\cite{GersEtal11}, Dictionary
learning~\citep{MairEtal09a, MairEtal09b}, deep
auto-encoders~\citep{HintSala06}, among others.  While complete
characterizations of each of these algorithms is beyond the scope of
the present manuscript, the general intuition driving these approaches
is to compute the $T$ by $I$ matrix, $\mathbf{X}$, that is closest
to the original $T$ by $J$ matrix, $\mathbf{Y}$, where (typically)
$I \ll J$.  The different approaches place different constraints on
what properties $\mathbf{X}$ must satisfy and which aspects of
the data are compared (and how) to characterize the match between
$\mathbf{X}$ and $\mathbf{Y}$.

Applying dimensionality reduction algorithms to $\mathbf{Y}$ yields a
$\mathbf{X}$ whose columns reflect weighted combinations (or nonlinear
transformations) of the original columns of $\mathbf{Y}$.  This has
two main consequences.  First, with each repeated dimensionality
reduction, the resulting $\mathbf{X}_n$ has lower and lower fidelity
(with respect to what the ``true'' $\mathbf{Y}_n$ might have looked
like without using dimensionality reduction to maintain scalability).
In other words, computing $\mathbf{X}_n$ is a lossy operation.
Second, whereas each columns of $\mathbf{Y}_n$ may always be mapped
directly onto specific pairs of columns of $\mathbf{Y}_{n-1}$, the
columns of $\mathbf{X}_n$ reflect weighted combinations and/or
nonlinear transformations of the columns of $\mathbf{Y}_n$.  Many
dimensionality reduction algorithms are invertible (or approximately
invertible).  However, attempting to map a given $\mathbf{X}_n$ back
onto the original feature space of $\mathbf{Y}_0$ will usually require
$\mathcal{O}(TK^{2n})$ space and therefore quickly becomes intractable
as $n$ or $K$ grow large.

@@

In general, graph measures take as input a matrix of
interactions (e.g., using the above notation, an $K$ by $K$
correlation matrix or binarized correlation matrix reconstituted from
a single timepoint's row of $\mathbf{Y}$) and return as output a set
of $K$ measures describing how each node (feature) sits within that
correlation matrix with respect to the rest of the population.  Widely
used measures include betweenness centrality~\citep[the proportion of
shortest paths between each pair of nodes in the population that
involves the given node in question; e.g., ][]{Newm05, OpsaEtal10,
  Bart04, GeisEtal08, Free77}; diversity and
dissimilarity~\citep[characterizations of how differently connected a
given node is from others in the population; e.g., ][]{Rao82, Lin09,
  RicoSzei06}; Eigenvector centrality and pagerank
centrality~\citep[measures of how influential a given node is within
the broader network; e.g., ][]{Newm08, Bona07, LohmEtal10,
  HaluEtal13}; transfer entropy and flow coefficients~\citep[a measure
of how much information is flowing from a given node to other nodes in
the network; e.g., ][]{HoneEtal07, Schr00}; $k$-coreness
centrality~\citep[a measure of the connectivity of a node within its
local sub-graph; e.g., ][]{AlvaEtal05, ChriFowl10}; within-module
degree~\citep[a measure of how many connections a node has to its
close neighbors in the network; e.g., ][]{RubiSpor10}; participation
coefficient~\citep[a measure of the diversity of a node's connections
to different sub-graphs in the network; e.g., ][]{RubiSpor10}; and
sub-graph centrality~\citep[a measure of a node's participation in all
of the network's sub-graphs; e.g., ][]{EstrRodr05}; among others.

@@

\paragraph*{Event.} We generated a fourth set of 10 synthetic datasets
whose underlying correlation matrices exhibited prolonged intervals of
stability, interspersed with abrupt changes.  For each dataset, we
used Equation~\ref{eqn:randcorr} to generate 5 random covariance
matrices.  We constructed a timeseries where each set of 60 consecutive samples
was drawn from a Gaussian with the same covariance matrix.  These
datasets were intended to simulate a system that undergoes occasional
abrupt state changes.

\subsubsection*{Functional neuroimaging data collected during story
  listening}
We examined an fMRI dataset collected by \cite{SimoEtal16} that the
authors have made publicly available at
\href{http://arks.princeton.edu/ark:/88435/dsp015d86p269k}{arks.princeton.edu/ark:/88435/dsp015d86p269k}.  The dataset
comprises neuroimaging data collected as participants listened
to an audio recording of a story (intact condition; 36 participants),
listened to time scrambled recordings of the same story (17
participants in the paragraph-scrambled condition listened to the
paragraphs in a randomized order and 36 in the word-scrambled
condition listened to the words in a randomized order), or lay resting
with their eyes open in the scanner (rest condition; 36
participants).  Full neuroimaging details may be found in the original
paper for which the data were collected~\citep{SimoEtal16}.

@@

\subsubsection*{Identifying robust decoding results}
The temporal decoding procedure we use to estimate which neural
features support ongoing cognitive processing is governed by many
parameters.  For example, Equation~\ref{eqn:timecorr} requires
defining a kernel function, which can take on different shapes and
widths.  For a fixed set of neural features, each of these parameters
can yield different decoding accuracies.  Further, the best decoding
accuracy for a given timepoint may be reliably achieved by one set of
parameters, whereas the best decoding accuracy for another timepoint
might be reliably achieved by a different set of parameters, and the
best decoding accuracy across \textit{all} timepoints might be
reliably achieved by still another different set of parameters.
Rather than attempting to maximize decoding accuracy, we sought to
discover the trends in the data that were robust to specific
classifier parameters choices.  Specifically, we sought to
characterize how decoding accuracy varied (under different
experimental conditions) as a function of which neural features were
considered.

@@

\section*{Results}
We sought to understand whether high-level cognition is supported by
dynamic patterns of high-order correlations.  To that end, we
developed a computational framework for estimating the dynamics of
high-order correlations in multivariate timeseries data (see
\textit{Dynamic inter-subject functional connectivity (DISFC)} and
\textit{Dynamic higher-order correlations}).  We evaluated the
efficacy of this framework at recovering known patterns in several
synthetic datasets (see \textit{Synthetic data}).  We then applied the
framework to a public fMRI dataset collected as participants listened
to an auditorally presented story, a temporally scrambled version of
the story, or underwent a resting state scan (see \textit{Functional
  neuroimaging data collected during story listening}).  We used the
relative decoding accuracies of classifiers trained on
different sets of neural features to estimate which types of features
reflected ongoing cognitive processing.

@@

\begin{figure}[tp]
  \centering
  \includegraphics[width=\textwidth]{figs/decode_level}
  \caption{\textbf{Across-participant decoding accuracy varies with
      correlation order and cognitive engagement.}
    \textbf{a. Decoding accuracy as a function of order: PCA.}
    \textit{Order} ($x$-axis) refers to the maximum order of dynamic
    correlations that were available to the classifiers (see
    \textit{Feature weighting and testing}).  The reported
    across-participant decoding accuracies are averaged over all
    kernel shapes and widths (see \textit{Identifying robust decoding
      results}).  The $y$-values are displayed relative to chance
    accuracy (intact: $\frac{1}{300}$; paragraph: $\frac{1}{272}$;
    word: $\frac{1}{300}$; rest: $\frac{1}{400}$).  The error bars
    denote 95\% confidence intervals across cross-validation folds
    (i.e., random assignments of participants to the training and test
    sets).  The colors denote the experimental condition.  Arrows
    denote sets of features that yielded reliably higher (upwards
    facing) or lower (downward facing) decoding accuracy than the mean
    of all other features (via a two-tailed test, thresholded at
    $p < 0.05$).  The circled values represent the maximum decoding
    accuracy within each experimental condition.
    \textbf{b. Normalized decoding accuracy as a function of order:
      PCA.}  This panel displays the same results as Panel a, but here
    each curve has been normalized to have a maximum value of 1 and
    and a minimum value of 0 (including the upper and lower bounds of
    the respective 95\% confidence intervals).  Panels a and b used
    PCA to project each high-dimensional pattern of dynamic
    correlations onto a lower-dimensional space.  \textbf{c.  Decoding
      accuracy as a function of order: eigenvector centrality.} This
    panel is in the same format as Panel a, but here eigenvector
    centrality has been used to project the high-dimensional patterns
    of dynamic correlations onto a lower-dimensional space.
    \textbf{d. Normalized decoding accuracy as a function of order:
      eigenvector centrality.} This panel is in the same format as
    Panel b, but here eigenvector centrality has been used to project
    the high-dimensional patterns of dynamic correlations onto a
    lower-dimensional space.}
  \label{fig:decoding}
\end{figure}

Both sets of temporal decoding analyses yielded qualitatively similar
results for the auditory (non-rest) conditions of the experiment
(Fig.~\ref{fig:decoding}: pink, yellow, and teal lines).  The highest
decoding accuracy for participants who listened to the intact
(unscrambled) story was achieved using high-order dynamic correlations
(PCA: second-order; eigenvector-centrality: fourth-order).  Scrambled
versions of the story were best decoded by lower-order correlations
(PCA/paragraph: first-order; PCA/word: order zero; eigenvector
centrality/paragraph: order zero; eigenvector centrality/word: order
zero).  The two sets of analyses yielded different decoding results on
resting state data (Fig.~\ref{fig:decoding}: purple lines).  We note
could be decoded only very slightly above chance; we speculate that
the decoders might have picked up on attentional drift, boredom, or
tiredness; we hypothesize these all increase throughout the resting
state scan.  Our decoders might be picking up on aspects of these
loosely defined cognitive states that are common across individuals.
The PCA-based approach achieved the highest resting state decoding
accuracy using order zero features (non-correlational
activation-based), whereas the eigenvector centrality-based approach
achieved the highest resting state decoding accuracy using
second-order correlations.  Taken together, these analyses indicate
that high-level cognitive processing (while listening to the intact
story) is reflected in the dynamics of high-order correlations in
brain activity, whereas lower-level cognitive processing (while
listening to scrambled versions of the story that lack rich meaning)
is reflected in the dynamics of lower-order correlations and
non-correlational activity dynamics.  Further, these patterns are
associated both with the underlying activity patterns (characterized
using PCA) and also with the changing relative positions that
different brain areas occupy in their associated networks
(characterized using eigenvector centrality).

@@

The notion that cognition is reflected in (and possibly mediated by)
patterns of network dynamics has been suggested by or proposed in
myriad empirical studies and reviews~\citep[e.g., ][]{DemeEtal19,
  Turk13, LuriEtal18, FongEtal19, ParkEtal18b, PretEtal17,
  RoyEtal19, LiegEtal19, ZouEtal19, ChanGlov10, GonzEtal19}.  Our
study extends this line of work by finding cognitively relevant
higher-order network dynamics that reflect ongoing cognition.  Our
findings complement other work that uses graph theory and topology to
characterize how brain networks reconfigure during
cognition~\citep[e.g., ][]{BassEtal06, ZhenEtal19, McInJirs19,
  TokeSomm19, SizeEtal18, ReimEtal17, BetzEtal19}.
